---
title: "ECHO autumn 2024 / Spring 2025 results"
format: 
  html:
    fig-width: 10
    fig-height: 10
    html-table-processing: none # to switch off automatic striping of table rows -- see https://github.com/quarto-dev/quarto-cli/issues/6945
execute:
  echo: false
---

```{r}
#| label: import
#| warning: false
#| message: false


# Import libraries
library("tidyverse")
library("lubridate")
library("glue")
library("here")
library("arrow")
library("gt")
library("scales")

# Import custom functions
source(here("analysis", "0-lib", "design.R"))

campaign_info <-
  campaign_info |>
  filter(
    campaign_start_date >= as.Date("2020-12-07"), # only select actual campaigns
  ) |>
  mutate(
    campaign_start_date_compact = format(campaign_start_date, "%Y%m%d"),
  ) |>
  # selection of campaigns for testing
  filter(
    campaign_start_date >= as.Date("2023-08-28"),
    campaign_start_date <= as.Date("2025-03-31"),
  )

# Import released data ----

#input_dir_snapshot <- here("output", "4-snapshot")
input_dir_snapshot <- here("release", "4-snapshot")

event_name <- c(
  "vax",
  "vax_alive",
  "covid_admitted",
  #"covid_admitted_primary",
  "covid_critcare",
  "covid_death"
)

milestone <- fct_inorder(c("Early", "Primary", "Final"))

eligibility_status <-
  rlang::quos(
    eligibility_status = case_when(
      group1=="age_above_eligiblity_threshold" & group1_value=="TRUE" ~ "age-eligible",
      group1=="clinical_priority" & group1_value=="TRUE" ~ "clinical-eligible",
      group1=="clinical_priority_only" & group1_value=="TRUE" ~ "clinical-eligible-only",
      group1=="any_eligibility" & group1_value=="FALSE" ~ "not eligible",
      .default = NA_character_
    )
  )

## import prior vaccine history ----

data_prior_vax <-
  campaign_info |>
  select(-age_date, -ends_with("_milestone_date"), -ends_with("_milestone_days"), -age_threshold) |>
  mutate(
    data = map(
      campaign_start_date_compact, \(.x)
      read_csv(
        fs::path(input_dir_snapshot, glue("report_snapshot_{.x}"), glue("prior_vax_table.csv")),
        col_types = c("cccciiiiiiidddddiiddddddddddddd")
      )
    )
  ) |>
  unnest(data) |>
  select(-starts_with("campaign_start")) |>
  mutate(
    !!!eligibility_status,
  )


## import vaccine cumulative incidence estimates ----

data_uptake <-
  campaign_info |>
  select(-age_date, -ends_with("_milestone_date"), -ends_with("_milestone_days"), -age_threshold) |>
  expand_grid(event_name, milestone) |>
  filter(
    event_name %in% c("vax", "vax_alive")
  ) |>
  mutate(
    data = pmap(
      list(campaign_start_date_compact, event_name, milestone), \(.x, .y, .z)
        read_csv(
          fs::path(input_dir_snapshot, glue("report_snapshot_{.x}"), glue("km_estimates_{.y}_milestones_{.z}.csv")),
          col_types = c("ccccidddDc")
        ) |>
        select(-milestone)
    )
  ) |>
  unnest(data) |>
  select(-starts_with("campaign_start")) |> 
  mutate(
    !!!eligibility_status,
  )

## import event IR and IRR estimates ----

data_contrasts <-
  campaign_info |>
  select(-age_date, -ends_with("_milestone_date"), -ends_with("_milestone_days"), -age_threshold) |>
  expand_grid(event_name) |>
  mutate(
    data = map2(
      campaign_start_date_compact, event_name, \(.x,.y)
        read_csv(
          fs::path(input_dir_snapshot, glue("report_snapshot_{.x}"), glue("contrasts_{.y}.csv")),
          col_types = c("cccciidldddddd") # "exposure" should be an integer col but some values exceed the int limit so coding as a double
        )
    )
  ) |>
  unnest(data) |>
  select(-starts_with("campaign_start")) |> 
  mutate(
    # TODO fix in the upstream script so they are already called groupx_value
    group1_value = label1,
    group2_value = label2
  ) |>
  mutate(
    !!!eligibility_status,
  ) |>
  mutate(
    # TODO find out why this isn't calculated in upstream code
    ir = n_event/exposure,
   # add confidence limits for incidence rate
    ir.ln.se = 1/sqrt(n_event),
    ir.low = exp(log(ir) + ir.ln.se*qnorm(0.025)),
    ir.high = exp(log(ir) + ir.ln.se*qnorm(0.975)),
    .after = "ir"
  ) |>
  mutate(
    irr = if_else(reference_row, 1, irr),
    irr.low = if_else(reference_row, 1, irr.low),
    irr.high = if_else(reference_row, 1, irr.high),
  )


## labelling functions for in-line text ----

lbl_n <- label_comma()
lbl_pct <- label_percent()
lbl_n_pct <- function(numer, denom) {
    glue("{lbl_n(numer)} ({lbl_pct(numer/denom)})")
}

```


# Results

We present results for each bi-yearly vaccination programme separately. 

Additional breakdowns for COVID-19 vaccine coverage and disease burden can be explored in an interactive dashboard, published online [link]

## Autumn 2024

```{r}
#| label: inline-cohort-autumn2024

select_campaign <- "Autumn 2024"

data_cohort_autumn2024 <- 
    data_prior_vax |>
    filter(campaign_label == select_campaign) 

n_total <- data_cohort_autumn2024 |> 
    filter(group1=="all", group2=="all") |> 
    pull(total)

n_age_over <- data_cohort_autumn2024 |> 
    filter(group1=="age_above_eligiblity_threshold", group1_value == "TRUE", group2=="all") |> 
    pull(total) |> sum()

n_age_under <- n_total - n_age_over

n_cp_only <- data_cohort_autumn2024 |> 
    filter(group1=="clinical_priority_only", group1_value == "TRUE", group2=="all") |> 
    pull(total)

n_eligible <- n_age_over + n_cp_only

n_eligible_both <- data_cohort_autumn2024 |> 
    filter(group1=="age_above_eligiblity_threshold", group1_value == "TRUE", group2=="primis_atrisk", group2_value == "TRUE") |> #change  to group2=="clinical_priority" once rerun
    pull(total) |>  sum()

```



### Cohort

A total of `r lbl_n(n_total)` people aged over 12 were alive and registered on 30 Septmber 2024, the Monday of the first week of the Autumn 2024 programme. Of those, `r lbl_n_pct(n_eligible, n_total)` were eligible for vaccination. `r lbl_n(n_age_over)` were eligible due to age (aged 65 years and over). Of those, `r lbl_n_pct(n_eligible_both, n_age_over)` were also eligible due to clinical vulnerability. Of the `r lbl_n(n_age_under)` people under the age threshold, `r lbl_n_pct(n_cp_only, n_age_under)` were eligible due clinical vulnerability. 

Table X shows characteristics of this cohort, by broad age groups.  

```{r}
#| label: table-cohort-autumn2024

data_cohort_autumn2024 |> 
    filter(group1 == "ageband4") |>
    select(
        ageband4 = group1_value, 
        group = group2, value = group2_value, n = total
    ) |>
    mutate(
        total = sum(n),
        prop = n / total,
        .by = c(ageband4, group)
    ) |>
    mutate(
      value = if_else(value=="All", NA_character_, value),
      prop = if_else(group=="all", NA_real_, prop)
    ) |>
    pivot_wider(
        id_cols = c(group, value),
        names_from = ageband4,
        values_from = c(n, prop),
        names_vary = "slowest",
        names_glue = "{ageband4}__{.value}"
    ) |>
    filter(value != "FALSE") |>
    mutate(
        row_shade = ((as.integer(fct_inorder(group)))%%2) == 0,
        value = if_else(value=="TRUE", "", value)
    ) |>
    mutate(
        group = if_else(row_number()==1, group, ""),
        .by = group
    ) |>  
    gt() |>
    tab_spanner_delim(delim = "__") |>
    fmt_percent(
        columns = ends_with("prop"),
        decimals = 1,
    ) |> 
    sub_missing(
        everything(),
        missing_text="-"
    ) |>
    cols_hide(c("row_shade")) |>
    tab_options(
      table.font.size = pct(60),
      data_row.padding = px(2),
      table.width = pct(100),
      ihtml.use_text_wrapping = FALSE
    ) |>
    tab_style(
      style = list(cell_fill(color = "gray96")),
      locations = cells_body(rows = row_shade)
    ) |> 
    tab_style(
      style = list(cell_fill(color = "white")),
      locations = list(cells_body(rows = !row_shade))
    )
```


### Coverage

```{r}

#| label: inline-cohort-autumn2024

data_uptake_autumn2024 <- 
    data_uptake |>
    filter(campaign_label == select_campaign) 

# age eligible cohort

cmlinc_primary_age_over <- data_uptake_autumn2024 |> 
    filter(event_name=="vax", milestone=="Primary", group1=="age_above_eligiblity_threshold", group1_value=="TRUE") |>
    summarise(
        total = cmlinc[group2=="all"],
        female = cmlinc[group2=="sex" & group2_value=="Female"],
        male = cmlinc[group2=="sex" & group2_value=="Male"],
        white = cmlinc[group2=="ethnicity5" & group2_value=="White"],
        black = cmlinc[group2=="ethnicity5" & group2_value=="Black or Black British"],
        deprivationmost = cmlinc[group2=="imd_quintile" & group2_value=="1 (most deprived)"],
        deprivationleast = cmlinc[group2=="imd_quintile" & group2_value=="5 (least deprived)"],
    )

cmlinc_final_age_over <- data_uptake_autumn2024 |> 
    filter(event_name=="vax", milestone=="Final", group1=="age_above_eligiblity_threshold", group1_value=="TRUE") |> 
    summarise(
        total = cmlinc[group2=="all"],
    )

# under age clinical priority groups

cmlinc_primary_ineligible <- data_uptake_autumn2024 |> 
    filter(event_name=="vax", milestone=="Primary", group1=="any_eligibility", group1_value=="FALSE") |>
    summarise(
        total = cmlinc[group2=="all"],
        female = cmlinc[group2=="sex" & group2_value=="Female"],
        male = cmlinc[group2=="sex" & group2_value=="Male"],
        white = cmlinc[group2=="ethnicity5" & group2_value=="White"],
        black = cmlinc[group2=="ethnicity5" & group2_value=="Black or Black British"],
        deprivationmost = cmlinc[group2=="imd_quintile" & group2_value=="1 (most deprived)"],
        deprivationleast = cmlinc[group2=="imd_quintile" & group2_value=="5 (least deprived)"],
    )

cmlinc_final_ineligible <- data_uptake_autumn2024 |> 
    filter(event_name=="vax", milestone=="Final",  group1=="any_eligibility", group1_value=="FALSE") |> 
    summarise(
        total = cmlinc[group2=="all"],
    )



# not eligible groups

cmlinc_primary_cp <- data_uptake_autumn2024 |> 
    filter(event_name=="vax", milestone=="Primary", group1=="clinical_priority_only", group1_value=="TRUE") |>
    summarise(
        total = cmlinc[group2=="all"],
        female = cmlinc[group2=="sex" & group2_value=="Female"],
        male = cmlinc[group2=="sex" & group2_value=="Male"],
        white = cmlinc[group2=="ethnicity5" & group2_value=="White"],
        black = cmlinc[group2=="ethnicity5" & group2_value=="Black or Black British"],
        deprivationmost = cmlinc[group2=="imd_quintile" & group2_value=="1 (most deprived)"],
        deprivationleast = cmlinc[group2=="imd_quintile" & group2_value=="5 (least deprived)"],
    )

cmlinc_final_cp <- data_uptake_autumn2024 |> 
    filter(event_name=="vax", milestone=="Final", group1=="clinical_priority_only", group1_value=="TRUE") |> 
    summarise(
        total = cmlinc[group2=="all"],
    )

```

Amongst the age-eligible cohort, the cumulative incidence of vaccination at 8 weeks was `r lbl_pct(cmlinc_primary_age_over$total)`. Vaccination incidence at 8 weeks was higher in women (`r lbl_pct(cmlinc_primary_age_over$female)`) versus men (`r lbl_pct(cmlinc_primary_age_over$male)`), higher in white (`r lbl_pct(cmlinc_primary_age_over$white)`) versus all other ethnicities, and in those living in areas that were least deprived (`r lbl_pct(cmlinc_primary_age_over$deprivationleast)`) versus most deprived (`r lbl_pct(cmlinc_primary_age_over$deprivationmost)`). Similar patterns remained up to 30 March 2025, the day before the start of the Spring 2025 programme, where the cumulative vaccination incidence of vaccination in age-eligible people was `r lbl_pct(cmlinc_final_age_over$total)`.

In those aged under 65 years who were eligible for vaccination due to clinical vulnerability, the cumulative incidence of vaccination at 8 weeks was `r lbl_pct(cmlinc_primary_cp$total)`. Vaccination incidence at 8 weeks was higher in women (`r lbl_pct(cmlinc_primary_cp$female)`) versus men (`r lbl_pct(cmlinc_primary_cp$male)`), higher in white (`r lbl_pct(cmlinc_primary_cp$white)`) versus all other ethnicities, and in those living in areas that were least deprived (`r lbl_pct(cmlinc_primary_cp$deprivationleast)`) versus most deprived (`r lbl_pct(cmlinc_primary_cp$deprivationmost)`). Similar patterns remained up to 30 March 2025, the day before the start of the Spring 2025 programme, where the cumulative vaccination incidence of vaccination in age-eligible people was `r lbl_pct(cmlinc_final_cp$total)`.

Amongst those under 65 years old who were not eligible for vaccination, the cumulative incidence of vaccination at 8 weeks was `r lbl_pct(cmlinc_primary_ineligible$total)`.  

Table X and Figure X shows vaccine uptake by eligibility group.



```{r}
#| label: table-uptake-autumn2024

data_contrasts_autumn2024 <- 
    data_contrasts |>
    filter(campaign_label == select_campaign) |>
    mutate(
      ir_rescaled = ir*(campaign_info |> filter(campaign_label==select_campaign) |> pull(final_milestone_days)),
    )

data_uptake_autumn2024 |> 
  filter(
    milestone=="Final", event_name=="vax",
    !is.na(eligibility_status)
  ) |>
  left_join(
    data_contrasts_autumn2024 |> select(-clinical_priority, -eligibility_status), # |> filter(!is.na(eligibility_status)), 
    by = c("campaign_label", "event_name", "group1", "group1_value", "group2", "group2_value")
  ) |>
  select(
    eligibility_status,
    group = group2, value = group2_value,
    n_obs, n_event,
    reference_row,
    cmlinc, cmlinc.low, cmlinc.high, irr, irr.low, irr.high
  ) |>
  mutate(
    group = fct_inorder(group),
    value = if_else(value=="All", NA_character_, value),
    cmlinc.ci = glue("{label_number(accuracy=0.1, scale=100)(cmlinc.low)} - {label_number(accuracy=0.1, scale=100)(cmlinc.high)}"),
    irr.high = if_else(irr.high>100, Inf, irr.high),
    irr.ci = glue("{label_number(accuracy=0.01, scale=1)(irr.low)} - {label_number(accuracy=0.01, scale=1)(irr.high)}"),
    irr.ci = if_else(reference_row, "-", irr.ci),
  ) |>
  arrange(group, eligibility_status) |>
  pivot_wider(
    id_cols = c(group, value),
    names_from = eligibility_status,
    values_from = c(cmlinc, cmlinc.ci, irr, irr.ci),
    names_vary = "slowest",
    names_glue = "{eligibility_status}__{.value}"
  ) |>
  mutate(
    row_shade = ((as.integer(fct_inorder(group)))%%2) == 0,
  ) |>
  mutate(
    group = if_else(row_number()==1, group, ""),
    .by = group
  ) |>     
  gt() |>
  tab_spanner_delim(delim = "__") |>
  # cols_merge(
  #   columns = c(cmlinc.low, cmlinc.high),
  #   pattern = "{1} - {2}"
  # ) |>
  fmt_percent(
    columns = ends_with(c("cmlinc")),
    decimals = 1,
  ) |> 
  fmt_number(
    columns = ends_with(c("irr")),
    decimals = 2
  ) |> 
  sub_missing(
    everything(),
    missing_text="-"
  ) |>
  cols_hide(c("row_shade")) |>
  tab_options(
      table.font.size = pct(60),
      data_row.padding = px(2),
      table.width = pct(100),
      ihtml.use_text_wrapping = FALSE
    ) |>
    tab_style(
      style = cell_fill(color = "gray96"),
      locations = cells_body(rows = row_shade)
    ) |> 
    tab_style(
      style = cell_fill(color = "white"),
      locations = cells_body(rows = !row_shade)
    )

```


```{r}
#| label: figure-uptake-autumn2024

data_uptake_autumn2024 |>
  filter(
    !is.na(eligibility_status),
    event_name=="vax",
    group2 != "ageband4"
  ) |>
    select(
        milestone,
        eligibility_status,
        group = group2, value = group2_value,
        cmlinc, cmlinc.low, cmlinc.high, 
    ) |>
  mutate(
    group = fct_inorder(group),
  ) |>
  arrange(milestone, group, eligibility_status) |>
  ggplot()+
  geom_pointrange(aes(y=value, x = cmlinc, xmin = cmlinc.low, xmax = cmlinc.high, colour = milestone), position = position_dodge(width=0.5))+
  facet_grid(rows=vars(group), cols=vars(eligibility_status), scales="free_y", space="free_y")+
  theme_bw()+
  theme(
    strip.text.y = element_text(angle = 0),
    strip.background = element_blank()
  )

```




### Disease

Table X and Figure X shows incidence of COVID-19-related disease by eligibility group. 


```{r}
#| label: table-burden-autumn2024

data_contrasts_autumn2024 <- 
  data_contrasts |>
  filter(campaign_label == select_campaign) |>
  mutate(
    ir_rescaled = ir*(campaign_info |> filter(campaign_label==select_campaign) |> pull(final_milestone_days)),
  )

  data_contrasts_autumn2024 |>
    filter(
      !is.na(eligibility_status),
      event_name %in% c("covid_admitted", "covid_critcare", "covid_death"),
    ) |>
    select(
      event_name,
      eligibility_status,
      group = group2, value = group2_value,
      n_obs, n_event,
      reference_row,
      ir, irr, irr.low, irr.high
    ) |>
    mutate(
      group = fct_inorder(group),
      value = if_else(value=="All", NA_character_, value),
      irr.high = if_else(irr.high>100, Inf, irr.high),
      irr.ci = glue("{label_number(accuracy=0.01, scale=1)(irr.low)} - {label_number(accuracy=0.01, scale=1)(irr.high)}"),
      irr.ci = if_else(reference_row, "-", irr.ci),
    ) |>
    arrange(group, eligibility_status) |>
    pivot_wider(
      id_cols = c(group, event_name, value),
      names_from = eligibility_status,
      values_from = c(n_obs, n_event, irr, irr.ci),
      names_vary = "slowest",
      names_glue = "{eligibility_status}__{.value}"
    ) |>
    mutate(
      row_shade = ((as.integer(fct_inorder(group)))%%2) == 0,
    ) |>
    mutate(
      group = if_else(row_number()==1, group, ""),
      .by = group
    ) |>
    gt(
      groupname_col = "event_name"
    ) |>
    tab_spanner_delim(delim = "__") |>
    # cols_merge(
    #   columns = c(cmlinc.low, cmlinc.high),
    #   pattern = "{1} - {2}"
    # ) |>
    fmt_number(
      columns = ends_with(c("irr")),
      decimal = 2
      ) |> 
    sub_missing(
      everything(),
      missing_text="-"
    ) |>
    cols_hide(c("row_shade")) |>
    tab_options(
      table.font.size = pct(60),
      data_row.padding = px(2),
      table.width = pct(100),
      ihtml.use_text_wrapping = FALSE
    ) |>
    tab_style(
      style = cell_fill(color = "gray96"),
      locations = cells_body(rows = row_shade)
    ) |> 
    tab_style(
      style = cell_fill(color = "white"),
      locations = cells_body(rows = !row_shade)
    )



```


```{r}
#| label: figure-burden-autumn2024

data_contrasts_autumn2024 |>
    filter(
      !is.na(eligibility_status),
      event_name %in% c("covid_admitted", "covid_critcare", "covid_death"),
    ) |>
    select(
      event_name,
      eligibility_status,
      group = group2, value = group2_value,
      n_obs, n_event,
      reference_row,
      ir, ir.low, ir.high,
      irr, irr.low, irr.high
    ) |>
  mutate(
    ir.high = if_else(is.nan(ir.high), Inf, ir.high),
    group = fct_inorder(group),
  ) |>
  arrange(event_name, eligibility_status, group) |>
  ggplot()+
  geom_pointrange(aes(y=value, x = ir, xmin = ir.low, xmax = ir.high, colour = eligibility_status), position = position_dodge(width=0.5))+
  facet_grid(rows=vars(group), cols=vars(event_name), scales="free", space="free_y")+
  scale_x_log10(label = label_number(scale=1000000*7), breaks = c(0.01,0.1,1,10,100,1000)/(1000000*7))+
  #coord_cartesian(xlim = c(0,0.00002))+
  labs(x = "Incidence rate, per 1M person-weeks")+
  theme_bw()+
  theme(
    strip.text.y = element_text(angle = 0),
    strip.background = element_blank()
  )


```






## Spring 2025

TODO (copy as for autumn 2024)

# Discussion

We present a new surveillance programme for COVID-19 vaccination and disease in England...


## Limitations

No documented eligibility does not mean that people were not eligible, for instance, we were not able to classify people as pregnant, working in a health or social care setting, or a close relative or contact of vulnerable people. There was also a degree of operational flexibility in how clinical vulnerability could be interpreted. Finally, clinical coding are imperfect, with likely underascertainment of conditions in the electronic record. 

## Future work

Continue for future covid-19 biyearly vaccination campaigns.

Extend to Influenza and RSV.

Full population coverage with EMIS

# References
